---
title: "Problem Set 5"
author: "Sydney Yeomans"
format:
  html:
    embed-resources: true
editor: visual
---

## Problem 1
Create a class to represent Wald-style normal approximation Confidence Intervals. Do this using S4

a. For the waldCI class, define the following:

```{r}

```

b. Use your waldCI class to create three objects:

```{r}

```

c. Show that your validator does not allow the creation of invalid confidence intervals: 

```{r}

```



## Problem 3
Repeat problem set 4, question 3, using plotly.

```{r}
install.packages("plotly")
library(plotly)
library(ggplot2)
library(tidyverse)
library(dplyr)
```

```{r}
#| echo: true
nyt <- read.csv("https://raw.githubusercontent.com/nytimes/covid-19-data/refs/heads/master/rolling-averages/us-states.csv")
```

a. How many major and minor spikes in cases were there?

```{r}
nyt$date <- as.Date(nyt$date)

nyt_spikes <- nyt %>%
              mutate(month = format(as.Date(date), "%Y-%m")) %>%  
              group_by(month) %>%
              summarize(monthly_cases = sum(cases, na.rm = TRUE)) %>%
              ungroup()

p <- plot_ly(nyt_spikes, x = ~ month, y = ~ monthly_cases, type = "scatter", 
        mode = "lines+markers")

p2 <- p %>% layout(title = "Monthly COVID Cases in the US",
                  xaxis = list(title="Month"),
                  yaxis = list(title="Total Cases"))
p2
```

From this we see the same thing we say in the last problem set, that there are 7 spikes (5 minor and 2 major). It is nice being able to hover over the dots and see how many cases there are thanks to plotly. 

b. For the states with the highest and lowest overall rates per population, what differences do you see in their trajectories over time?

```{r}

```

c. Identify, to the best of your ability without a formal test, the first five states to experience Covid in a substantial way.

```{r}

```




