---
title: "Problem Set 5"
author: "Sydney Yeomans"
format:
  html:
    embed-resources: true
editor: visual
---

## Problem 1
Create a class to represent Wald-style normal approximation Confidence Intervals. Do this using S4

a. For the waldCI class, define the following:

```{r}

```

b. Use your waldCI class to create three objects:

```{r}

```

c. Show that your validator does not allow the creation of invalid confidence intervals: 

```{r}

```



## Problem 3
Repeat problem set 4, question 3, using plotly.

```{r}
install.packages("plotly")
library(plotly)
library(ggplot2)
library(tidyverse)
library(dplyr)
```

```{r}
#| echo: true
nyt <- read.csv("https://raw.githubusercontent.com/nytimes/covid-19-data/refs/heads/master/rolling-averages/us-states.csv")
```

a. How many major and minor spikes in cases were there?

```{r}
#| echo: true
nyt$date <- as.Date(nyt$date)

nyt_spikes <- nyt %>%
              mutate(month = format(as.Date(date), "%Y-%m")) %>%  
              group_by(month) %>%
              summarize(monthly_cases = sum(cases, na.rm = TRUE)) %>%
              ungroup()

p <- plot_ly(nyt_spikes, x = ~ month, y = ~ monthly_cases, type = "scatter", 
        mode = "lines+markers")

p2 <- p %>% layout(title = "Monthly COVID Cases in the US",
                  xaxis = list(title="Month"),
                  yaxis = list(title="Total Cases"))
p2
```

From this we see the same thing we say in the last problem set, that there are 7 spikes (5 minor and 2 major). It is nice being able to hover over the dots and see how many cases there are thanks to plotly. 

b. For the states with the highest and lowest overall rates per population, what differences do you see in their trajectories over time?

```{r}
high_low <- nyt %>% 
            group_by(state) %>%
            summarize(highest_rate = max(cases_avg_per_100k, na.rm = TRUE)) %>%
            filter(highest_rate == max(highest_rate) | highest_rate == min(highest_rate))%>%
            ungroup()

nyt_highlow <- nyt %>%
               filter(state == "American Samoa" | state == "Maine")

nyt_highlow$date <- as.Date(nyt_highlow$date)

q <- plot_ly(nyt_highlow, x = ~ date, y = ~ cases_avg_per_100k, color = ~ state, 
             type = "scatter", mode = "lines")

q2 <- q %>% layout(title = "Comparison of States with Highest and Lowest Overall COVID Cases",
                  xaxis = list(title="Date"),
                  yaxis = list(title="Cases on Average per 100K"))
q2

```

c. Identify, to the best of your ability without a formal test, the first five states to experience Covid in a substantial way.

```{r}
nyt_first <- nyt %>%
             group_by(state) %>%
             mutate(date = as.Date(date),
                    year = format(date, "%Y"),
                    state = factor(state)) %>%
             filter(year == "2020")

first_four_mon <- nyt %>%
                  group_by(state) %>%
                  mutate(date = as.Date(date),
                         year = format(date, "%Y"),
                         state = factor(state)) %>%
                  filter(date >= as.Date("2020-03-01"),
                         date <= as.Date("2020-04-30"))

r <- plot_ly(first_four_mon, x = ~ date, y = ~ cases_avg_per_100k, color = ~ state, 
             type = "scatter", mode = "lines")

r2 <- r %>% layout(title = "First States to Experience COVID in a Substantial Way (2020)",
                  xaxis = list(title="Date"),
                  yaxis = list(title="Cases on Average per 100K"))
r2

```

From this first plot, that focuses solely on the first four months of 2020, we see that New York, New Jersey, and Louisiana are way above the other states so we can guess that they are some of the first states to experience COVID in a substantial way. The other two are convoluted but we can just take the top 7-8 and then go from there. Zooming in on the plotly we already made we can see other states that have early peaks are Massachusetts, Connecticut, Michigan, and DC (which were the ones showing up last time with ggplot as well). So let us subset the data to just these states.

```{r}
seven_states <- c("New York", "New Jersey", "Louisiana", "Massachusetts", 
                  "Connecticut", "District of Columbia", "Michigan")

top_sev <- nyt %>%
            filter(state %in% seven_states) %>%
            mutate(date = as.Date(date),
                   year = format(date, "%Y"),
                   state = factor(state, levels = seven_states)) %>%
            filter(date >= as.Date("2020-03-01"),
                   date <= as.Date("2020-04-30"))

s <- plot_ly(top_sev, x = ~ date, y = ~ cases_avg_per_100k, color = ~ state, 
             type = "scatter", mode = "lines")

s2 <- s %>% layout(title = "First Seven States to Experience COVID in a Substantial Way (2020)",
                  xaxis = list(title="Date"),
                  yaxis = list(title="Cases on Average per 100K"))
s2

```

If we zoom in and try to see which of these seven reached 10 cases on average per 100k we can pick out the top 5. Of course NY, NJ, and LA, show up again and if we focus on this threshold the other two are MA and CT. Now I will make a last graph just focusing on these 5: NY, NJ, MA, LA, and CT.

```{r}
funf_state <- c("New York", "New Jersey", "Louisiana", "Massachusetts", 
                  "Connecticut")

top_five <- nyt %>%
            filter(state %in% funf_state) %>%
            mutate(date = as.Date(date),
                   year = format(date, "%Y"),
                   state = factor(state, levels = funf_state)) %>%
            filter(year == "2020")

t <- plot_ly(top_five, x = ~ date, y = ~ cases_avg_per_100k, color = ~ state, 
             type = "scatter", mode = "lines")

t2 <- t %>% layout(title = "First 5 States to Experience COVID in a Substantial Way (2020)",
                  xaxis = list(title="Date"),
                  yaxis = list(title="Cases on Average per 100K"))
t2

```


